A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-04 paper.Rmd
finished revising section 5.6

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the fitted value distribution on the power of conventional and visual tests. For conventional tests, only the power curves of appropriate tests which are RESET tests and Breusch-Pagan tests for non-linearity pattern and heteroskedasticity pattern respectively are shown. For visual tests, note that more evaluations on lineup with uniform fitted value distribution were collected, so to have a fair comparison, we randomly sample five from the total eleven evaluations to estimate the power curves. Re-sampling produces the multiple curves for the uniform, and provides an indication of the variation of the power.

In terms of visual tests, we do not observe significant difference between the power curve of the uniform distribution and the power curve of the discrete uniform distribution. Normal and lognormal distributions produce lower power than uniform and discrete uniform distributions in regards of residual departure patterns, but the gap is larger when the residual plot contains heteroskedasticity pattern. This suggests that residual departures displayed with evenly distributed fitted values are visually more attractive and effective. Besides, the skewness of the fitted value distribution plays a role in the degree of presence of visual features as lognormal distribution produce the lowest power for both non-linearity and heteroskedasticity patterns. A symmetric fitted value distribution can better reveal the underlying shape.

Although it is unusual to consider fitted value distribution or predictor distribution in power analysis, it is not so surprised that the power of conventional tests vary because of fitted value distribution. For instance, the power of the RESET test which is effectively a F test depends on the realizations of the additional predictors included in the auxiliary regression equation [@jamshidian2007study; @olvera2019relationship; @zhang2018practical]. And those additional predictors are linear transformation of random vectors following certain predictor distributions. What is truly unexpected is the huge variability in power of conventional tests compared to visual tests. The discrete uniform distribution constantly produce the highest power while the normal distribution and the lognormal distribution produce the lowest power for non-linearity and heteroskedasticity pattern respectively. The difference in power corresponding to different distributions peak at around 90% when the effect size is at a moderate level. This huge difference is undesirable as is not uncommon for analysts to omit the factor of predictor distribution in power calculation and sample size calculation.

>>>>

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the fitted value distribution on the power of conventional (top row) and visual (bottom row) tests for both the non-linearity (left column) and heteroskedasticity (right column) patterns. For conventional tests, only the power curves of appropriate tests which are RESET tests and Breusch-Pagan tests for non-linearity pattern and heteroskedasticity pattern respectively are shown. For visual tests, note that, more evaluations on lineup with uniform fitted value distribution were collected, so to have a fair comparison, we randomly sample five from the total eleven evaluations to estimate the power curves. Re-sampling produces the multiple curves for the uniform, and provides an indication of the variation of the power.

Perhaps surprisingly, the visual tests have more consistent power across the different fitted value distributions. For the non-linear pattern, there is almost no power difference. For the heteroskedastic pattern, uniform and discrete have higher power than normal and lognormal. The likely reason is that these latter two have less observations in the tails where the heteroskedastic pattern needs to be detected.

<!-- In terms of visual tests, we do not observe significant difference between the power curve of the uniform distribution and the power curve of the discrete uniform distribution. Normal and lognormal distributions produce lower power than uniform and discrete uniform distributions in regards of residual departure patterns, but the gap is larger when the residual plot contains heteroskedasticity pattern. This suggests that residual departures displayed with evenly distributed fitted values are visually more attractive and effective. Besides, the skewness of the fitted value distribution plays a role in the degree of presence of visual features as lognormal distribution produce the lowest power for both non-linearity and heteroskedasticity patterns. A symmetric fitted value distribution can better reveal the underlying shape.-->

The variation in power in the conventional tests is at first sight, shocking. However, it is discussed, albeit rarely, in the testing literature. See, for example, @jamshidian2007study, @olvera2019relationship and  @zhang2018practical which show derivations and use simulation to assess the effect of the observed distribution of the predictors on test power. The big differences in power curves that are seen in Figure \ref{fig:different-x-dist-poly-power} echoes the results reported in these articles.

<!-- Although it is unusual to consider fitted value distribution or predictor distribution in power analysis, it is not so surprised that the power of conventional tests vary because of fitted value distribution. For instance, the power of the RESET test which is effectively a F test depends on the realizations of the additional predictors included in the auxiliary regression equation [@jamshidian2007study; @olvera2019relationship; @zhang2018practical]. And those additional predictors are linear transformation of random vectors following certain predictor distributions. What is truly unexpected is the huge variability in power of conventional tests compared to visual tests. The discrete uniform distribution constantly produce the highest power while the normal distribution and the lognormal distribution produce the lowest power for non-linearity and heteroskedasticity pattern respectively. The difference in power corresponding to different distributions peak at around 90% when the effect size is at a moderate level. This huge difference is undesirable as is not uncommon for analysts to omit the factor of predictor distribution in power calculation and sample size calculation. -->

### 2023-04-04 paper.tex
finished revising section 5.6

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the
fitted value distribution on the power of conventional and visual tests.
For conventional tests, only the power curves of appropriate tests which
are RESET tests and Breusch-Pagan tests for non-linearity pattern and
heteroskedasticity pattern respectively are shown. For visual tests,
note that more evaluations on lineup with uniform fitted value
distribution were collected, so to have a fair comparison, we randomly
sample five from the total eleven evaluations to estimate the power
curves. Re-sampling produces the multiple curves for the uniform, and
provides an indication of the variation of the power.

In terms of visual tests, we do not observe significant difference
between the power curve of the uniform distribution and the power curve
of the discrete uniform distribution. Normal and lognormal distributions
produce lower power than uniform and discrete uniform distributions in
regards of residual departure patterns, but the gap is larger when the
residual plot contains heteroskedasticity pattern. This suggests that
residual departures displayed with evenly distributed fitted values are
visually more attractive and effective. Besides, the skewness of the
fitted value distribution plays a role in the degree of presence of
visual features as lognormal distribution produce the lowest power for
both non-linearity and heteroskedasticity patterns. A symmetric fitted
value distribution can better reveal the underlying shape.

Although it is unusual to consider fitted value distribution or
predictor distribution in power analysis, it is not so surprised that
the power of conventional tests vary because of fitted value
distribution. For instance, the power of the RESET test which is
effectively a F test depends on the realizations of the additional
predictors included in the auxiliary regression equation
\citep{jamshidian2007study, olvera2019relationship, zhang2018practical}.
And those additional predictors are linear transformation of random
vectors following certain predictor distributions. What is truly
unexpected is the huge variability in power of conventional tests
compared to visual tests. The discrete uniform distribution constantly
produce the highest power while the normal distribution and the
lognormal distribution produce the lowest power for non-linearity and
heteroskedasticity pattern respectively. The difference in power
corresponding to different distributions peak at around 90\% when the
effect size is at a moderate level. This huge difference is undesirable
as is not uncommon for analysts to omit the factor of predictor
distribution in power calculation and sample size calculation.

>>>>

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the
fitted value distribution on the power of conventional (top row) and
visual (bottom row) tests for both the non-linearity (left column) and
heteroskedasticity (right column) patterns. For conventional tests, only
the power curves of appropriate tests which are RESET tests and
Breusch-Pagan tests for non-linearity pattern and heteroskedasticity
pattern respectively are shown. For visual tests, note that, more
evaluations on lineup with uniform fitted value distribution were
collected, so to have a fair comparison, we randomly sample five from
the total eleven evaluations to estimate the power curves. Re-sampling
produces the multiple curves for the uniform, and provides an indication
of the variation of the power.

Perhaps surprisingly, the visual tests have more consistent power across
the different fitted value distributions. For the non-linear pattern,
there is almost no power difference. For the heteroskedastic pattern,
uniform and discrete have higher power than normal and lognormal. The
likely reason is that these latter two have less observations in the
tails where the heteroskedastic pattern needs to be detected.

The variation in power in the conventional tests is at first sight,
shocking. However, it is discussed, albeit rarely, in the testing
literature. See, for example, \citet{jamshidian2007study},
\citet{olvera2019relationship} and \citet{zhang2018practical} which show
derivations and use simulation to assess the effect of the observed
distribution of the predictors on test power. The big differences in
power curves that are seen in Figure
\ref{fig:different-x-dist-poly-power} echoes the results reported in
these articles.
